---
title: "STAT 5361 -  HW 7"
author: Patrick Toman^[<patrick.toman@uconn.edu>; Ph.D. student at Department of Statistics,
  University of Connecticut.]
date: "`r format(Sys.time(), '%d %B %Y')`"
output: pdf_document
header-includes:
  - \usepackage[ruled,vlined,linesnumbered]{algorithm2e}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = F)
```


# Problem 5.3.1 

No idea if any of this is right. 

## Part (i)

## Part (ii) 

### Pseudo-code

\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\KwIn{$n =$ sample size,$\theta$ = user defined scale parameter}
\KwResult{$n$ Simulated Observations from density $g(x)$}
\BlankLine
\For{i in 1:n}{
    Allocate empty vector $X \in \mathcal{R}^{n \times 1}$ \;
    Draw $U \sim Bernoulli\left(p = \frac{2 \Gamma(\theta)}{2 \Gamma(\theta) + \Gamma(\theta+1/2)}\right)$ \;
    \eIf{$U = 1$}{
        $X[i] \sim Gamma(\theta,1)$\;
    }{
        $X[i] \sim Gamma(\theta+\frac{1}{2},1)$\;
    }
}
\end{algorithm}

### Implementation


```{r,gamma_mix_sim}
rm(list=ls())


theta <- 3.25
p <- 2*gamma(theta)/(2*gamma(theta) + gamma(theta+0.5))

x = seq(0.1,15,by=0.1)


my_gamma_mix_sim <- function(size,theta){
  
  X <- rep(NA,size)
  
  p <- 2*gamma(theta)/(2*gamma(theta) + gamma(theta+0.5))
  
  for(i in 1:size){
    
    U <- rbinom(1,1,p)
    
    if(U == 1){
      
      X[i] <- rgamma(1,shape = theta,rate=1)
      
    }else{
      
      X[i] <- rgamma(1,shape = theta+1/2,rate=1)
      
    }
    
  }
  
  return(X)
  
  
  
}


set.seed(1022)


my_simulated_gammamix <- my_gamma_mix_sim(size = 10000,theta=theta)




simulated_gammamix <- bmixture::rmixgamma(n=10000,weight = c(p,1-p),alpha = c(theta,theta+1/2),beta = rep(1,2)) 
density_gammamix <- bmixture::dmixgamma(x=x,weight = c(p,1-p),alpha = c(theta,theta+1/2),beta = rep(1,2)) 



hist(simulated_gammamix,prob=T,nclass = 30,col = 'red')
hist(my_simulated_gammamix,prob=T,nclass = 30,add=T,col = 'blue')
lines(x,density_gammamix,type = 'l',lwd=3)

```



## Part (iii) - Rejection Sampling

### Pseudo-code

### Implementation

```{r,densities}
gy <- function(y,theta){(2*y^(theta-1) + y^(theta-1/2))*exp(-y)}

fy <- function(y,theta){sqrt(4+y)*y^(theta-1)*exp(-y)}

gy_density <- gy(y=x,theta=theta)
fy_density <- fy(y=seq(0,20,length.out = 10000),theta=theta)

plot(gy_density,type='l',col='blue')
lines(fy_density,type = 'l',col='red')

curve(sqrt(4+x)*x^(theta-1)*exp(-x),from = 0.001,to=15)

```

```{r,rejection_sampling}


rejection_sampler <- function(n,theta,alpha){
  
  X <- rep(NA,n)
  
  p <- 2*gamma(theta)/(2*gamma(theta) + gamma(theta+0.5))
  
  accepted_ct <- 0 
  
  rejected_ct <- 0 
  
  i <- 1 
  
  while(accepted_ct < n){
    
    y <- my_gamma_mix_sim(size=1,theta=theta)
    
    gy <- gy(y=y,theta=theta)
    
    fy_set <- fy(y=y,theta=theta)
    
    U <- runif(1,0,1)
    
    if(U < fy_set/(alpha*gy)){
      
      X[i] <- y
      
      i <- i + 1
      
      accepted_ct <- accepted_ct + 1
      
    }else{
      
      rejected_ct <- rejected_ct + 1
      
    }
  
  }
  
  return(list('Sample'=X,'RejectCt'=rejected_ct))
}

```

```{r,rejection_sample_test}

n <- 10000

rejection <- rejection_sampler(n=n,theta = theta,alpha=3)

plot_df <- cbind.data.frame('Length'=seq(0,20,length.out = 10000),'RejectionSample'=rejection$Sample,'Target'=fy_density)

plot(density(rejection$Sample))


plot_df %>% 
  ggplot()+
  geom_density(aes(x=RejectionSample))+
  ylab('Density') + xlab('y') + 
  ggtitle('Estimated Density')+
  theme_minimal() -> plt_estimated_density

plot_df %>% 
  ggplot()+
  geom_line(aes(x=Length,y=Target))+
  theme_minimal() + ylab('f(y)') + xlab('y') +
  ggtitle('Target Density (Un-normalized)') -> plt_true_density

gridExtra::grid.arrange(plt_estimated_density,plt_true_density, ncol=2)

```

# Problem 6.3.1 

$\pi(\mu_1),\pi(\mu_2)\sim N(0,\tau^2 = 10^2)$ and $\pi(1/\sigma^2_1) \sim \Gamma(a,b)$


```{r}

delta <- 0.7 # true value to be estimated based on the data
n <- 100
set.seed(123)
u <- rbinom(n, prob = delta, size = 1)
x <- rnorm(n, ifelse(u == 1, 7, 10), 0.5)

mylike <- function(delta, x) {
    prod(delta * dnorm(x, 7, 0.5) + (1 - delta) * dnorm(x, 10, 0.5))
}

```